---
title: "Chapter 6"
author: "Silja Tammi"
date: "12/8/2023"
output: html_document
---

# Chapter 6

```{r, include=FALSE}
# Load libraries
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lme4)
```

## 1. Meet and Repeat: PART I using the RATS data

RATS dataset contains data from a nutrition study conducted in three groups of rats (Crowder and Hand, 1990).  The three groups were put on different diets, and each animal’s body weight (grams) was recorded repeatedly (approximately weekly, except in week seven when two recordings were taken) over a 9-week period. The question of most interest is whether the growth profiles of the three groups differ.

RATS data has been converted into long format in meet_and_repeat.R, so I import the data:

```{r}
RATSL <- read.table("./data/RATSL.txt", header = T)

# convert ID and Group into factors
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

# Draw ggplot with `Time` on the x-axis and `Weight` on the y-axis
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=20)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))

```

From the plot we can see that the starting weight differs a lot for the three groups of rats. In all groups, weight increases during the study time, more so in groups 2 and 3. There is also one rat in group 2 that differs from the other rats in the group.

Next, we inspect the tracking to see if rats that have higher weight values at the beginning tend to have higher values throughout the study. 

The tracking phenomenon can be seen more clearly in a plot of the standardized values of each
observation, i.e., the values obtained by subtracting the relevant occasion mean from the original observation and then dividing by the corresponding visit standard deviation:

$$standardised(x) = \frac{x - mean(x)}{ sd(x)}$$
```{r}
# Group by WD and standardise the variable Weight
RATSL <- RATSL %>%
  group_by(WD) %>%
  mutate(stdweight = scale(Weight)) %>%
  ungroup()

# Plot again with the standardised Weight
ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=20)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized weight")
```

Let'sproduce graphs showing average (mean) profiles for each group along with some indication of the variation of the observations at each time point, in this case the standard error of mean

$$se = \frac{sd(x)}{\sqrt{n}}$$

```{r}
# Summary data with mean and standard error of Weight by Group and Time 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight) ) %>%
  ungroup()

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.4)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")

```

We can see that the measurements are not independent. The mean of group 1 differs from that of groups 2 and 3, whereas the standard errors of groups 2 and 3 overlap. Variation of weight in rat groups 2 and 3 is bigger than in group 1 (there seemed to be outliers in group 2 and 3 in the plot in previous step increasing variation). Increase in weight seems to be only slightly bigger in group 2.

Next let's check for these outliers. 

```{r}
# Create a summary data by Group and ID with mean as the summary variable
RATSL8S <- RATSL %>%
  group_by(Group, ID) %>%
  summarise(mean=mean(Weight) ) %>%
  ungroup()

ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Days 1-64") +
  ggtitle("Mean weight in the groups, all rats included") +
  theme(plot.title = element_text(hjust = 0.5))
```

There seems to be a clear outlier in group 2. Maybe the rats with lower mean weights in groups 1 and 3 could also be considered outliers, but here I only consider the rat in group 2 as an outlier and remove it from the data.

```{r}
# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
RATSL8S1 <- RATSL8S[-12, ]

ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Days 1-64") +
  ggtitle("Mean weight in the groups, outlier in group 2 removed") +
  theme(plot.title = element_text(hjust = 0.5))
```


Graphical presentations indicated a lack of difference in the three diets, but let's perform a t-test to check for difference. 

Baseline measurements of the outcome variable in a longitudinal study are often correlated with the chosen summary measure and using such measures in the analysis can often lead to substantial gains in precision when used appropriately as a covariate in an analysis of covariance. We can illustrate the analysis on the data using the weight corresponding to time zero taken prior to the start of study as the baseline covariate. 


```{r, warning=F}

# Add the baseline from the original data as a new variable to the summary data
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')

RATSL8S2 <- RATSL8S %>%
 mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSL8S2)
summary(fit)
confint(fit)

# Compute the analysis of variance table for the fitted model with anova()
# anova(fit)
anova(fit)

```

Summary for the linear model shows that the baseline of the groups is associated with weight  and that the association is statistically significant (p-value 5.02e-08), whereas the group (=diet) is not associated with weight, although there is slight indication of association for group 2 (p-value 0.0888).The 95% confidence intervals are also wide and include the zero, so there is no evidence that the group differences differ from zero.

Anova confirms the lack of evidence for a group difference (p-value 0.07586). 


## 2. Meet and Repeat: PART II using the BPRS data

BPRS data consists of 40 male subjects randomly assigned to one of two treatment groups and each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0) and then at weekly intervals for eight weeks. The BPRS assesses the level of 18 symptom constructs such as hostility, suspiciousness, hallucinations and grandiosity; each of these is rated from one 
(not present) to seven (extremely severe). The scale is used to evaluate patients suspected of having schizophrenia.

BPRS data has been converted into long format in meet_and_repeat.R, so I import the data:

```{r}
BPRSL <- read.table("./data/BPRSL.txt", header = T)

# convert ID and Group into factors
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)


# Draw the plot
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

```

The bprs values seem to decrease in both treatment groups during the treatment weeks 1-8, but there doesn't seem to be significant differences between the groups.

Let's fit a multiple linear regression model to see if this is indeed correct.

```{r}
# fit a multiple linear regression model with `pbrs` as response and `week` and `treatment` as explanatory variables

# create a regression model RATS_reg
BPRSL_reg <- lm(data=BPRSL, bprs ~ week + treatment)

# print out a summary of the model
summary(BPRSL_reg)
```

In the summary we can see that time (week) is statistically significantly (p-value <2e-16) negatively correlated with pbrs value, so during treatment weeks 1-8 the bprs value goes down. However, the type of treatment (treatment) is not statistically significant, so the type of treatment doesn't have an effect on the decrease of pbrs value, as could be inspected from the plot above. 

The lm model assumes independence of the repeated measures of bprs, and this assumption is highly unlikely. So, now we will move on to consider both some more appropriate graphics and appropriate models.

To begin the more formal analysis of the bprs data, we will first fit the *random intercept model* for the same two explanatory variables: `week` and `treatment`. Fitting a random intercept model allows the linear regression fit for each study subject to differ in *intercept* from other subjects. So what we do is we combine the variance components and the single level regression model and we get a random intercept model.

```{r}
# Create a random intercept model as subject as the random effect
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)
```

We see here again that treatment and time don't correlate with each other. The t value for treatment is close to zero (0.532), so it's p-value is large, whereas the t-statistic for week is far from zero (-10.896) with corresponding small p-value, which indicates that it is unlikely to get such results by chance and that the null hypothesis of no correlation between time and bprs should not be rejected.

Now we can move on to fit the *random intercept and random slope model* to the BPRS data. Fitting a random intercept and random slope model allows the linear regression fits for each individual to differ in intercept but also in slope. This way it is possible to account for the individual differences in the bprs profiles, but also the effect of time.

```{r}
# create a random intercept and random slope model with `week` and `subject` as the random effects
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# Compute the ANOVA analysis of variance tables of the models 
anova(BPRS_ref1, BPRS_ref)
```

BPRS_ref1 (a random intercept and random slope model) is a better fit than BPRS_ref (a random intercept model) in terms of chi-squared statistics (7.2721) and p-value of the likelihood ratio test (0.02636).

Finally, we can fit a random intercept and slope model that allows for a treatment group × time interaction, even though we saw that treatment and time don't correlate with each other.

```{r}
# Write the same model as in the previous exercise but add `week` * `treatment` interaction.
BPRS_ref2 <- lmer(bprs ~ week + treatment + (week | subject) + (week * treatment), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)
```

As could be expected, the model with the week * treatment interaction variable is not statistically better than the model without interaction (chi-squared statistic 3.1712, p-value 0.07495). So here the addition of the interaction was not necessary and didn't improve the model.